---
title: "Cannabis sativa SimpleTidy_GeneCoEx"
author: "Karen Cristine Goncalves"
date: "`r format(Sys.time(), '%A %d %B %Y')`"
format: 
  html:
    toc: true
    embed-resources: true
    default-image-extension: svg
    df-print: kable
editor: source
---

## Initial data overview

```{r, include=F, echo=F, warning=F, eval=T}
suppressMessages(library(tidyverse))
```


```{r, include=F, eval=T, echo=F, results='asis'}
metadata <- read_delim("metadata/metadata_pca.txt", 
                       col_names = T, delim = "\t",
                       show_col_types = F) 
```

```{r, include=T, eval=T, echo=F, results='asis'}
nDataSets = nrow(metadata)
bioprojects = unique(metadata$BioProject) %>% length
tissues = unique(metadata$tissue) %>% length

cat("
We found a total of", nDataSets, "RNAseq datasets for _Cannabis sativa_ available in NCBI, belonging to", bioprojects, "different projects,", tissues, "different tissues/organs, and sequenced from both single- and paired-end libraries (with variable read sizes, from 35 to 150).\n
    ")
```
::: {layout-ncol="3"}
```{r, include=T, eval=T, echo=F, results='asis'}
#| label: tbl-projects
# samples_per_project <- 
table(metadata$BioProject) %>%
  as.data.frame() %>%
  rename(BioProject = Var1,
         `Number of Samples` = Freq) 
```

```{r, include=T, eval=T, echo=F, results='asis'}
#| label: tbl-tissues
# samples_per_tissue <- 
table(metadata$tissue) %>%
  as.data.frame() %>%
  rename(Tissue = Var1,
         `Number of Samples` = Freq) 
```

```{r, include=T, eval=T, echo=F, results='asis'}
#| label: tbl-layouts

# samples_per_layout <- 
  table(metadata$Layout) %>%
  as.data.frame() %>%
  rename(`Sequencing layout` = Var1,
         `Number of Samples` = Freq) 
```
:::


Read trimming and filtering (fastp) was repeated to include the shorter-read samples:

-   for samples with original average read length \< 40 bases, the minimum size allowed after trimming was 30 bases
-   for samples with original average read length \> 40 bases, the minimum size allowed after trimming was 50 bases.

## Read mapping, counting and TPM computation

Reads were aligned to _C. sativa_’s female genome (downloaded from Ensembl Plants release 61: [genome](https://ftp.ensemblgenomes.ebi.ac.uk/pub/plants/release-59/fasta/cannabis_sativa_female/dna/Cannabis_sativa_female.cs10.dna.toplevel.fa.gz) and [GTF annotation](https://ftp.ensemblgenomes.ebi.ac.uk/pub/plants/release-61/gtf/cannabis_sativa_female/Cannabis_sativa_female.cs10.61.gtf.gz)) using STAR.

Given the difference in average read lengths between the datasets, the genome indexing step was taking using four different values for the `--sjdbOverhang` parameter: 30, 70, 90 and 144.

The option `--quantMode` was added to STAR to obtain read counts. Strandness was infered using RSeQC `infer_experiment.py` script, alignment BAM files and the genome annotation in BED format.

Subsequently, read counts (for each replicate, according to their strandness) were combined into a count matrix using R 4.4.0 and the tidyverse package.

For TPM computation, the GTF file was used to compute gene lengths (through exon overlap) (using R packages rtracklayer and GenomicRanges).

## PCA and gene filtering


### PCA

Principal component analysis of all available RNAseq datasets in NCBI, based on gene expression levels (log~10~(TPM+1)), after removal of non-expressed genes (no reads detected in any sample of the dataset). Since no clear issues were visible, the analysis proceeded with all datasets.


:::{layout-ncol="2"}
::: {#fig-wholeData_tissue }
![](plots/PCA06_tissue.svg){.lightbox width="100%"}
:::

:::{#fig-wholeData_project}

![](plots/PCA06_BioProject.svg){.lightbox width="100%"}

:::

:::

### Filtering

```{r, include=T, eval=T, echo=F, results='asis'}
ngenes = read_delim("counts/TPM.tsv", progress = F, show_col_types = F) %>% nrow()
ngenes_kept = read_csv("results/Filtered_TPM.csv", progress = F, show_col_types = F) %>% nrow()
filtered = ngenes - ngenes_kept
proportion = (ngenes_kept*100)/ngenes 
tmp = paste0("were kept (", signif(proportion, digits = 4), "% of all genes).")

cat("After filtering weakly expressed genes,", filtered, 
    "were removed from the dataset and", ngenes_kept, 
    tmp, "
    ")
```

### PCA post gene filtering

No obvious difference was observed in the PCA before and after gene filtering, indicating that most of the genes removed from the dataset had TPM = 0 in all samples (as those genes were removed prior to the initial PCA).

:::{#fig-filteredData}
![](plots/PCA_filtered_data.svg){.lightbox width="75%"}
:::

